---
title: "NHS Waiting List and Service Performance Analysis"
author: "Asif Khandokar"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: hide
---

# 0. Project Overview

## 0.1 Project Aim

This project analyses NHS England referral-to-treatment (RTT)
waiting-list data for three London NHS trusts between April 2024
and March 2026.

The aim is to examine changes in waiting-list size and waiting
times, compare selected specialties, and identify patterns
that warrant further investigation.

## 0.2 Research Questions

1. How have waiting-list sizes changed over the study period?
2. How have the percentages of pathways within 18 weeks and
   over 52 weeks changed?
3. Which selected specialties show increasing waiting lists
   or persistent long waits?
4. What differences between trusts warrant further investigation?

## 0.3 Project Scope

The selected trusts are:

- Barts Health NHS Trust (R1H).
- Imperial College Healthcare NHS Trust (RYJ).
- Guy's and St Thomas' NHS Foundation Trust (RJ1).

The selected specialties are:

- General Surgery Service (C_100).
- Trauma and Orthopaedic Service (C_110).
- Ophthalmology Service (C_130).

The data describe incomplete RTT pathways at the end of each
reporting month. Pathways are not necessarily unique patients,
because one patient can have more than one pathway.

# 1. Background and NHS Context

## 1.1 What Is Referral to Treatment?

Referral to Treatment (RTT) measures the time from referral
to the start of consultant-led treatment for non-emergency care.

An incomplete RTT pathway represents a case where treatment
has not yet started at the end of the reporting month.
The patient may be waiting for a diagnostic test, a specialist
appointment or a procedure.

A patient can have more than one pathway for different
conditions. Therefore, pathway counts should not be interpreted
as counts of unique patients.

## 1.2 Why Analyse Waiting Lists?

Waiting-list analysis helps service managers understand
changes in the number of pathways awaiting treatment and
the length of time people have been waiting.

Examining trends by trust and specialty can identify areas
that need further investigation. However, waiting-list data
alone cannot establish why performance has changed.

## 1.3 Measures Used in This Project

This project will examine:

- Total incomplete pathways: the waiting-list size at month-end.
- Percentage within 18 weeks: the proportion of incomplete
  pathways with a wait of no more than 18 weeks.
- Pathways waiting over 52 weeks: the number and proportion
  experiencing long waits.
- Median waiting time: the middle waiting time among
  incomplete pathways, as published in the dataset.
- Monthly and annual changes: movements in waiting-list
  size and waiting-time indicators.

## 1.4 Interpreting Comparisons

Trusts differ in size, services and the complexity of the
patients they treat. A larger waiting list does not, by itself,
indicate poorer performance.

This project compares the same selected specialties across
trusts and examines both counts and percentages. Differences
will be described as patterns requiring investigation rather
than evidence of their underlying causes.

Monthly waiting lists are snapshots. Adding their totals
across months would not give the number of distinct patients
or pathways seen during the study period.

## 1.5 Source

NHS England:
[Referral to Treatment Waiting Times — Data Collection](https://digital.nhs.uk/data-and-information/data-collections-and-data-sets/data-collections/consultant-led-referral-to-treatment-waiting-times-rtt).

# 2. Data Source and Definitions

## 2.1 Data Source

The data come from NHS England's monthly Consultant-led
Referral to Treatment (RTT) Waiting Times publications.

This project uses the "Provider" sheet from 24 monthly
"Incomplete Provider" Excel workbooks covering April 2024
to March 2026.

The source publication pages are:

- [RTT Data 2024–25](https://www.england.nhs.uk/statistics/statistical-work-areas/rtt-waiting-times/rtt-data-2024-25/)
- [RTT Data 2025–26](https://www.england.nhs.uk/statistics/statistical-work-areas/rtt-waiting-times/rtt-data-2025-26/)

Original workbooks are retained unchanged in the data folder.
Their filenames identify the reporting month and, where
indicated, whether the file is revised.

## 2.2 Selected Trusts and Specialties

The analysis focuses on three London NHS trusts:

| Provider code | NHS trust |
|:--------------|:----------|
| R1H | Barts Health NHS Trust |
| RYJ | Imperial College Healthcare NHS Trust |
| RJ1 | Guy's and St Thomas' NHS Foundation Trust |

Three specialties are selected:

| Treatment function code | Specialty |
|:------------------------|:----------|
| C_100 | General Surgery Service |
| C_110 | Trauma and Orthopaedic Service |
| C_130 | Ophthalmology Service |

These specialties are recorded for all three trusts throughout
the selected period. Reporting coverage and missing values
will also be checked within the R workflow.

## 2.3 Unit of Analysis

Each observation in the prepared specialty dataset will
represent one trust, one specialty and one reporting month.

The expected dataset contains:

24 months × 3 trusts × 3 specialties = 216 observations.

These are aggregated observations, rather than individual
patient records.

## 2.4 Variable Dictionary

| Prepared variable | Meaning | Type |
|:------------------|:--------|:-----|
| reporting_month | Reporting month, stored as its first day; measures describe month-end | Date |
| provider_code | Identifier for the NHS trust | Character |
| provider_name | Name of the NHS trust | Character |
| specialty_code | Published treatment function code | Character |
| specialty_name | Published treatment function name | Character |
| total_pathways | Total incomplete pathways at month-end | Numeric count |
| within_18_weeks | Number of incomplete pathways within 18 weeks | Numeric count |
| pct_within_18_weeks | Published proportion of incomplete pathways within 18 weeks | Numeric proportion (0–1) |
| over_52_weeks | Published "Total 52 plus weeks" count | Numeric count |
| pct_over_52_weeks | Proportion calculated as over_52_weeks / total_pathways | Numeric proportion (0–1) |
| median_wait_weeks | Published median waiting time in weeks, converted to numeric during cleaning | Numeric |
| source_file | Original workbook filename | Character |

These shorter names are assigned during data preparation.
Proportions are multiplied by 100 when displayed as percentages
in tables and charts.

## 2.5 Percentage Representation

Excel stores the percentage measures as proportions.

For example, a stored value of 0.65 represents 65%.

We will retain proportions for calculations and format them
as percentages in tables and charts. A change from 60% to
65% is an increase of 5 percentage points.

## 2.6 Workbook Structure

In the March 2026 Provider sheet, the column headings appear
on row 14 and the data begin on row 15.

The workbooks contain weekly waiting-time bands alongside
published summary measures. Column headings vary slightly
across the collected files, so the import process will check
and standardise them.

Each trust also has a "Total" row with treatment function
code C_999. These rows will be kept separate from specialty
records to prevent double-counting.

The sum of our three selected specialties will be labelled
as a selected-specialty total, rather than the whole trust's
waiting list.

## 2.7 Initial Data Quality Checks

Before analysis, the workflow will check:

- All 24 reporting months are present.
- Each trust–specialty–month combination is unique.
- Required measures are numeric and available.
- Pathway counts are non-negative whole numbers.
- Counts within 18 weeks and over 52 weeks do not exceed
  total pathways.
- Published proportions agree with their corresponding
  counts and totals, allowing for numerical precision.
- Missing values remain distinguishable from zero.

The presence of a record does not guarantee that it is
complete or accurate. Any issues will be investigated and
documented before interpreting the results.

# 3. R Setup and Data Import

## 3.1 Load Packages

Packages provide functions for importing Excel files,
preparing data and creating visualisations.

```{r load-packages}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## 3.2 Locate the Monthly Workbooks

The original Excel workbooks are stored in the project's
data folder. We search for Incomplete Provider files,
including any stored in subfolders.

```{r locate-files}
# Find the monthly Provider workbooks.
rtt_files <- list.files(
  path = "data",
  pattern = "^Incomplete-Provider-.*\\.xlsx$",
  full.names = TRUE,
  recursive = TRUE,
  ignore.case = TRUE
)

# Check the number of files found.
length(rtt_files)

# Display their filenames.
basename(rtt_files)
```

The expected number of workbooks is 24. Finding 24 files
is an initial check; we will verify their reporting months
after importing them.

## 3.3 Import One Workbook

We first import a single workbook to understand its structure
before combining all 24 months.

```{r import-sample}
# Locate the March 2026 workbook.
sample_file <- rtt_files[
  grepl("Incomplete-Provider-Mar26-", basename(rtt_files))
]

# Confirm exactly one matching file.
stopifnot(length(sample_file) == 1)

# Display the workbook's sheet names.
excel_sheets(sample_file)

# Import the Provider sheet.
# Skip the first 13 rows so row 14 becomes the column headings.
rtt_sample <- read_excel(
  path = sample_file,
  sheet = "Provider",
  skip = 13
)
```

## 3.4 Inspect the Imported Data

We examine the dimensions, column names and first few records
to check whether the import has worked as expected.

```{r inspect-sample}
# Number of rows and columns.
dim(rtt_sample)

# Original column names.
names(rtt_sample)

# Display the first six records using the identifier columns.
head(
  rtt_sample %>%
    select(
      `Provider Code`,
      `Provider Name`,
      `Treatment Function Code`,
      `Treatment Function`
    )
)
```

## 3.5 Select the Project Sample

We use provider and specialty codes to select the records
relevant to the project. Codes provide more consistent
identifiers than organisation names.

```{r select-sample}
# Define the three trusts.
selected_trusts <- c("R1H", "RYJ", "RJ1")

# Define the three specialties.
selected_specialties <- c("C_100", "C_110", "C_130")

# Select the relevant records and give columns shorter names.
rtt_sample_selected <- rtt_sample %>%
  filter(
    `Provider Code` %in% selected_trusts,
    `Treatment Function Code` %in% selected_specialties
  ) %>%
  transmute(
    provider_code = `Provider Code`,
    provider_name = `Provider Name`,
    specialty_code = `Treatment Function Code`,
    specialty_name = `Treatment Function`,
    total_pathways = `Total number of incomplete pathways`,
    within_18_weeks = `Total within 18 weeks`,
    pct_within_18_weeks = `% within 18 weeks`,
    over_52_weeks = `Total 52 plus weeks`,
    pct_over_52_weeks = `% 52 plus weeks`,
    median_wait_weeks = `Average (median) waiting time (in weeks)`
  )

# Check the number of selected records.
nrow(rtt_sample_selected)

# Display a compact table.
knitr::kable(
  rtt_sample_selected %>%
    select(
      provider_code,
      specialty_name,
      total_pathways,
      within_18_weeks,
      over_52_weeks
    ),
  caption = "Selected trust and specialty records: March 2026"
)
```

## 3.6 Create a Monthly Import Function

A function lets us apply the same preparation steps to every
workbook without copying the code 24 times.

We read the reporting period from cell C5, rather than relying
only on the filename. Dates are stored as the first day of
the reporting month for sorting and plotting; the measures
describe the month-end waiting list.

```{r monthly-import-function}
read_rtt_month <- function(file_path) {

  # Read the reporting period from the workbook.
  period_text <- read_excel(
    file_path,
    sheet = "Provider",
    range = "C5",
    col_names = FALSE
  )[[1]][1]

  # Convert the month name and year into a Date.
  period_parts <- strsplit(trimws(period_text), "\\s+")[[1]]
  month_number <- match(period_parts[1], month.name)
  year_number <- as.integer(period_parts[2])

  if (is.na(month_number) || is.na(year_number)) {
    stop("Could not identify reporting month: ", basename(file_path))
  }

  reporting_month <- as.Date(
    sprintf("%04d-%02d-01", year_number, month_number)
  )

  # Read the monthly Provider data.
  monthly_data <- read_excel(
    file_path,
    sheet = "Provider",
    skip = 13
  )

  # Select the project records and standardise column names.
  selected_data <- monthly_data %>%
    filter(
      `Provider Code` %in% selected_trusts,
      `Treatment Function Code` %in% selected_specialties
    ) %>%
    transmute(
      reporting_month = reporting_month,
      provider_code = `Provider Code`,
      provider_name = `Provider Name`,
      specialty_code = `Treatment Function Code`,
      specialty_name = `Treatment Function`,
      total_pathways = `Total number of incomplete pathways`,
      within_18_weeks = `Total within 18 weeks`,
      pct_within_18_weeks = `% within 18 weeks`,
      over_52_weeks = `Total 52 plus weeks`,
      median_wait_weeks = `Average (median) waiting time (in weeks)`,
      source_file = basename(file_path)
    ) %>%
    mutate(
      pct_over_52_weeks = over_52_weeks /
        na_if(total_pathways, 0)
    )

  return(selected_data)
}
```

## 3.7 Combine All Monthly Files

The lapply() function runs our import function once for each
workbook. bind_rows() combines the resulting tables.

```{r combine-months}
# Import each workbook into a list of monthly tables.
rtt_monthly_list <- lapply(rtt_files, read_rtt_month)

# Combine and sort the records.
rtt_data <- bind_rows(rtt_monthly_list) %>%
  arrange(reporting_month, provider_code, specialty_code)

# Check the combined dataset.
nrow(rtt_data)
n_distinct(rtt_data$reporting_month)
range(rtt_data$reporting_month)

# Check the number of records in each month.
knitr::kable(
  rtt_data %>% count(reporting_month, name = "records"),
  caption = "Number of selected records per reporting month"
)
```

# 4. Initial Data Inspection

## 4.1 Inspect Dimensions and Variable Types

We check the size and structure of the combined dataset.
Dates, identifiers and numerical measures should have
appropriate types before analysis.

```{r inspect-combined-data}
# Number of rows and columns.
dim(rtt_data)

# Display each variable's type and sample values.
glimpse(rtt_data)
```

## 4.2 Summarise Numerical Measures

Summary statistics provide an initial view of the ranges
and distributions. They can also reveal missing values
or unexpected measurements.

```{r summarise-measures}
rtt_data %>%
  select(
    total_pathways,
    within_18_weeks,
    pct_within_18_weeks,
    over_52_weeks,
    pct_over_52_weeks,
    median_wait_weeks
  ) %>%
  summary()
```

## 4.3 Check Coverage by Trust and Specialty

Each trust–specialty combination should have 24 records,
one for each reporting month. Duplicate and missing-month
checks will follow in Section 5.

```{r inspect-group-coverage}
knitr::kable(
  rtt_data %>%
    count(provider_code, specialty_code, specialty_name,
          name = "records"),
  caption = "Record counts by trust and specialty"
)
```

# 5. Data Cleaning and Quality Assurance

## 5.1 Correct the Median Waiting-Time Type

The median waiting-time column was imported as text.
We convert it to numeric and check whether any non-missing
values fail conversion.

```{r convert-median-wait}
# Retain the original values for checking.
median_wait_original <- rtt_data$median_wait_weeks

# Convert text values to numbers.
median_wait_numeric <- as.numeric(
  trimws(as.character(median_wait_original))
)

# Identify values that could not be converted.
conversion_issues <- rtt_data %>%
  filter(
    !is.na(median_wait_original),
    is.na(median_wait_numeric)
  ) %>%
  select(
    reporting_month,
    provider_code,
    specialty_code,
    median_wait_weeks,
    source_file
  )

# Display any conversion issues.
nrow(conversion_issues)
conversion_issues

# Stop if conversion would lose an existing value.
stopifnot(nrow(conversion_issues) == 0)

# Apply the checked conversion.
rtt_data$median_wait_weeks <- median_wait_numeric

# Confirm the corrected type and numerical summary.
class(rtt_data$median_wait_weeks)
summary(rtt_data$median_wait_weeks)
```

## 5.2 Check Duplicate Records

Each trust–specialty–month combination should appear once.
We flag repeated combinations for investigation rather
than deleting them automatically.

```{r check-duplicates}
duplicate_records <- rtt_data %>%
  count(reporting_month, provider_code, specialty_code) %>%
  filter(n > 1)
```

## 5.3 Check Missing Combinations

We compare the observed records with all expected
combinations of 24 months, three trusts and three specialties.

```{r check-missing-combinations}
expected_records <- expand_grid(
  reporting_month = seq(
    as.Date("2024-04-01"),
    as.Date("2026-03-01"),
    by = "month"
  ),
  provider_code = selected_trusts,
  specialty_code = selected_specialties
)

missing_records <- expected_records %>%
  anti_join(
    rtt_data,
    by = c("reporting_month", "provider_code", "specialty_code")
  )

unexpected_records <- rtt_data %>%
  anti_join(
    expected_records,
    by = c("reporting_month", "provider_code", "specialty_code")
  )
```

## 5.4 Check Missing Values

Missing values are counted for every variable. They are
not replaced with zero because missing and zero have
different meanings.

```{r check-missing-values}
missing_values <- data.frame(
  variable = names(rtt_data),
  missing_count = as.integer(colSums(is.na(rtt_data)))
)

knitr::kable(
  missing_values,
  caption = "Missing values by variable"
)
```

## 5.5 Summarise Record Coverage Checks

```{r coverage-check-summary}
coverage_checks <- data.frame(
  check = c(
    "Duplicate trust–specialty–month combinations",
    "Missing expected combinations",
    "Records outside the project scope"
  ),
  issues_found = c(
    nrow(duplicate_records),
    nrow(missing_records),
    nrow(unexpected_records)
  )
)

knitr::kable(
  coverage_checks,
  caption = "Record coverage checks"
)
```


## 5.6 Validate Numerical Measures

Counts should be non-negative whole numbers. Waiting-time
proportions should lie between zero and one, and subgroup
counts should not exceed total pathways.

We also compare the published proportion within 18 weeks
with the proportion calculated from its count and total.

```{r validate-measures}
# Small tolerance for numerical precision.
tolerance <- 1e-8

# Identify records with invalid numerical measures.
invalid_records <- rtt_data %>%
  filter(
    !is.finite(total_pathways) |
    !is.finite(within_18_weeks) |
    !is.finite(over_52_weeks) |
    !is.finite(pct_within_18_weeks) |
    !is.finite(pct_over_52_weeks) |
    !is.finite(median_wait_weeks) |
    total_pathways < 0 |
    within_18_weeks < 0 |
    over_52_weeks < 0 |
    total_pathways != floor(total_pathways) |
    within_18_weeks != floor(within_18_weeks) |
    over_52_weeks != floor(over_52_weeks) |
    within_18_weeks > total_pathways |
    over_52_weeks > total_pathways |
    within_18_weeks + over_52_weeks > total_pathways |
    pct_within_18_weeks < 0 |
    pct_within_18_weeks > 1 |
    pct_over_52_weeks < 0 |
    pct_over_52_weeks > 1 |
    median_wait_weeks < 0
  )

# Compare the published proportion with an independent calculation.
percentage_mismatches <- rtt_data %>%
  mutate(
    calculated_proportion = within_18_weeks /
      na_if(total_pathways, 0),
    difference = abs(
      pct_within_18_weeks - calculated_proportion
    )
  ) %>%
  filter(
    is.na(calculated_proportion) |
    difference > tolerance
  )

# Display the validation results.
numerical_checks <- data.frame(
  check = c(
    "Records with invalid numerical measures",
    "Published 18-week proportion mismatches"
  ),
  issues_found = c(
    nrow(invalid_records),
    nrow(percentage_mismatches)
  )
)

knitr::kable(
  numerical_checks,
  caption = "Numerical validation checks"
)
```

```{r enforce-quality-checks}
stopifnot(
  nrow(duplicate_records) == 0,
  nrow(missing_records) == 0,
  nrow(unexpected_records) == 0,
  all(missing_values$missing_count == 0),
  nrow(invalid_records) == 0,
  nrow(percentage_mismatches) == 0
)
```

# 6. Calculating Performance Indicators

## 6.1 Calculate Monthly and Annual Changes

Changes are calculated separately for each trust and specialty,
with records ordered by reporting month.

Waiting-list changes are expressed as counts and percentages.
Changes in waiting-time proportions are expressed in
percentage points.

```{r calculate-changes}
rtt_analysis <- rtt_data %>%
  group_by(provider_code, specialty_code) %>%
  arrange(reporting_month, .by_group = TRUE) %>%
  mutate(
    # Change from the preceding month.
    monthly_change_pathways =
      total_pathways - lag(total_pathways),

    monthly_change_pct =
      100 * monthly_change_pathways /
      na_if(lag(total_pathways), 0),

    monthly_change_18wk_pp =
      100 * (
        pct_within_18_weeks -
        lag(pct_within_18_weeks)
      ),

    # Change from the same month one year earlier.
    annual_change_pathways =
      total_pathways - lag(total_pathways, 12),

    annual_change_pct =
      100 * annual_change_pathways /
      na_if(lag(total_pathways, 12), 0),

    annual_change_18wk_pp =
      100 * (
        pct_within_18_weeks -
        lag(pct_within_18_weeks, 12)
      ),

    annual_change_52wk_pp =
      100 * (
        pct_over_52_weeks -
        lag(pct_over_52_weeks, 12)
      )
  ) %>%
  ungroup() %>%
  arrange(reporting_month, provider_code, specialty_code)
```

The first month of each series has no monthly comparison.
The first 12 months have no annual comparison. These missing
change values are expected because earlier observations
are outside the study period.

Using a 12-row lag represents one year here because our
coverage checks confirmed one record per consecutive month
for each trust–specialty combination.

## 6.2 Inspect the Latest Annual Changes

The table compares March 2026 with March 2025.

```{r latest-annual-changes}
latest_changes <- rtt_analysis %>%
  filter(reporting_month == max(reporting_month)) %>%
  select(
    provider_code,
    specialty_name,
    total_pathways,
    annual_change_pathways,
    annual_change_pct,
    annual_change_18wk_pp,
    annual_change_52wk_pp
  )

knitr::kable(
  latest_changes,
  digits = 2,
  col.names = c(
    "Trust",
    "Specialty",
    "Pathways",
    "Annual change",
    "Annual change (%)",
    "Within 18 weeks change (pp)",
    "Over 52 weeks change (pp)"
  ),
  caption = "March 2026 compared with March 2025"
)
```

# 7. Exploratory Analysis and Visualisation

## 7.1 How Have Waiting-List Sizes Changed?

We plot monthly incomplete pathways by trust, with a separate
panel for each specialty. Counts represent month-end snapshots.

```{r waiting-list-trends, fig.width=10, fig.height=7}
# Create readable trust labels for charts.
trust_labels <- c(
  "R1H" = "Barts Health",
  "RJ1" = "Guy's and St Thomas'",
  "RYJ" = "Imperial College Healthcare"
)

ggplot(
  rtt_analysis,
  aes(
    x = reporting_month,
    y = total_pathways,
    colour = provider_code,
    group = provider_code
  )
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.4) +
  facet_wrap(
    ~ specialty_name,
    ncol = 1,
    labeller = label_wrap_gen(width = 40)
  ) +
  scale_colour_manual(
    values = c(
      "R1H" = "#0072B2",
      "RJ1" = "#D55E00",
      "RYJ" = "#009E73"
    ),
    breaks = names(trust_labels),
    labels = unname(trust_labels)
  ) +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y"
  ) +
  scale_y_continuous(
    labels = function(x) {
      format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
    }
  ) +
  labs(
    title = "Waiting-list trends by trust and specialty",
    subtitle = "April 2024–March 2026",
    x = NULL,
    y = "Incomplete RTT pathways",
    colour = NULL,
    caption = paste(
      "Source: NHS England monthly RTT Provider workbooks.",
      "Pathways are not unique patients."
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

All panels use the same vertical scale to support comparison
of waiting-list sizes. Differences in trust size and service
mix should be considered when interpreting the counts.

### Interpretation

Barts Health recorded growth in General Surgery and Trauma
and Orthopaedics waiting lists during much of 2025, followed
by reductions towards March 2026. Imperial College Healthcare
showed an overall decline across the three selected specialties.

At Guy's and St Thomas', the General Surgery waiting list
declined during 2024 before increasing during 2025. Its large
March 2025–March 2026 percentage increase should therefore
be considered alongside the relatively low starting count.

These trends describe changes in month-end waiting-list
sizes. They do not establish whether changes resulted from
referral demand, treatment activity, administrative changes
or other factors.


## 7.2 How Has the Proportion Within 18 Weeks Changed?

```{r within-18-week-trends, fig.width=10, fig.height=7}
ggplot(
  rtt_analysis,
  aes(
    x = reporting_month,
    y = pct_within_18_weeks,
    colour = provider_code,
    group = provider_code
  )
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.4) +
  facet_wrap(
    ~ specialty_name,
    ncol = 1,
    labeller = label_wrap_gen(width = 40)
  ) +
  scale_colour_manual(
    values = c(
      "R1H" = "#0072B2",
      "RJ1" = "#D55E00",
      "RYJ" = "#009E73"
    ),
    breaks = names(trust_labels),
    labels = unname(trust_labels)
  ) +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y"
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = function(x) paste0(round(x * 100), "%")
  ) +
  labs(
    title = "Proportion of waiting pathways within 18 weeks",
    subtitle = "April 2024–March 2026",
    x = NULL,
    y = "Incomplete pathways within 18 weeks (%)",
    colour = NULL,
    caption = "Source: NHS England monthly RTT Provider workbooks."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

Higher values indicate a larger share of the month-end
waiting list within 18 weeks. This is not the percentage
of patients treated within 18 weeks during that month.

### Interpretation

General Surgery shows different patterns across the trusts.
Imperial's proportion within 18 weeks increased over the
study period. Guy's and St Thomas' improved into early 2025,
then declined, with a marked fall around October 2025.
This change warrants investigation alongside its growing
waiting list; the chart alone cannot establish the cause.

In Ophthalmology, Guy's and St Thomas' and Imperial ended
the period with higher proportions within 18 weeks than
at the start. Barts improved into 2025 before declining
towards the end of the period.

In Trauma and Orthopaedics, Guy's and St Thomas' and
Imperial generally improved over the full period.
Barts fluctuated and finished below the other two trusts.

The count and proportion charts should be read together.
Waiting-list growth can occur alongside an increase in
the share within 18 weeks, as seen in Barts' annual
General Surgery and Orthopaedics comparisons.

## 7.3 How Has the Proportion Waiting Over 52 Weeks Changed?

We examine long waits as a share of each specialty's
month-end waiting list. A falling proportion does not
necessarily mean the number of long-wait pathways fell,
so counts will also be considered when interpreting results.

```{r over-52-week-trends, fig.width=10, fig.height=7}
ggplot(
  rtt_analysis,
  aes(
    x = reporting_month,
    y = pct_over_52_weeks,
    colour = provider_code,
    group = provider_code
  )
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.4) +
  facet_wrap(
    ~ specialty_name,
    ncol = 1,
    labeller = label_wrap_gen(width = 40)
  ) +
  scale_colour_manual(
    values = c(
      "R1H" = "#0072B2",
      "RJ1" = "#D55E00",
      "RYJ" = "#009E73"
    ),
    breaks = names(trust_labels),
    labels = unname(trust_labels)
  ) +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y"
  ) +
  scale_y_continuous(
    labels = function(x) paste0(round(x * 100, 1), "%"),
    expand = expansion(mult = c(0, 0.05))
  ) +
  expand_limits(y = 0) +
  labs(
    title = "Proportion of waiting pathways over 52 weeks",
    subtitle = "April 2024–March 2026",
    x = NULL,
    y = "Incomplete pathways over 52 weeks (%)",
    colour = NULL,
    caption = "Source: NHS England monthly RTT Provider workbooks."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

Lower values indicate a smaller share of pathways
experiencing waits over 52 weeks. All panels use the
same vertical scale.

### Interpretation

Imperial's General Surgery long-wait proportion fell from
approximately 16% in April 2024 to below 1% by March 2026.
Barts also ended the period substantially lower, although
its proportion increased during part of 2025 before falling.

In Trauma and Orthopaedics, all three trusts ended with
lower long-wait proportions than at the start. Temporary
increases show that the improvement was not continuous.

Ophthalmology long-wait proportions remained comparatively
low and approached zero towards the end of the period.
The shared vertical scale makes smaller differences in
this specialty less visible.

These results describe proportions. Changes in the number
of long-wait pathways must also be examined to distinguish
a falling count from an increase in the total waiting list.

# 8. Trust and Specialty Comparisons

## 8.1 Compare the Latest Reporting Month

We compare trusts within each specialty using waiting-list
counts, waiting-time proportions and median waiting times.

```{r latest-performance-table}
latest_performance <- rtt_analysis %>%
  filter(reporting_month == max(reporting_month)) %>%
  arrange(specialty_code, provider_code) %>%
  transmute(
    Trust = unname(trust_labels[provider_code]),
    Specialty = specialty_name,
    Pathways = total_pathways,
    `Within 18 weeks (%)` = 100 * pct_within_18_weeks,
    `Over 52 weeks (count)` = over_52_weeks,
    `Over 52 weeks (%)` = 100 * pct_over_52_weeks,
    `Median wait (weeks)` = median_wait_weeks
  )

knitr::kable(
  latest_performance,
  digits = 2,
  caption = "Selected specialty performance: March 2026"
)
```

The comparisons are descriptive and are not adjusted for
patient complexity or differences in service provision.
They do not represent whole-trust performance.

### Interpretation

In March 2026, ophthalmology had the highest proportion of pathways
within 18 weeks and the shortest median waiting time at each of the
three selected trusts.

For trauma and orthopaedics, Barts Health had the largest waiting list
(8,564 pathways), the lowest proportion within 18 weeks (55.73%) and
the longest median wait (15.83 weeks). However, Guy’s and St Thomas’
had the highest number and proportion waiting over 52 weeks:
80 pathways, representing 1.64% of its specialty waiting list.

General surgery results were relatively close for the proportion
within 18 weeks, ranging from 55.28% to 58.34%. The proportion waiting
over 52 weeks varied more, from 0.10% at Imperial College Healthcare
to 1.36% at Barts Health.

These measures describe different parts of the waiting-time
distribution. A shorter median wait does not necessarily mean fewer
very long waits. Comparisons are descriptive and do not account for
differences in case complexity, referral patterns or service provision.

## 8.2 Changes in the Number of Long-Wait Pathways

The previous chart showed the proportion of pathways waiting over
52 weeks. This comparison examines the underlying counts between
March 2025 and March 2026, alongside changes in those proportions.

```{r annual-long-wait-comparison}
# Calculate annual changes within each trust and specialty.
long_wait_changes <- rtt_analysis %>%
  group_by(provider_code, specialty_code) %>%
  arrange(reporting_month, .by_group = TRUE) %>%
  mutate(
    previous_year_over_52 = lag(over_52_weeks, 12),
    annual_change_over_52 = over_52_weeks - previous_year_over_52
  ) %>%
  ungroup()

# Display the latest month and its comparison with one year earlier.
long_wait_comparison <- long_wait_changes %>%
  filter(reporting_month == max(reporting_month)) %>%
  arrange(specialty_code, provider_code) %>%
  transmute(
    Trust = unname(trust_labels[provider_code]),
    Specialty = specialty_name,
    `Over 52 weeks: March 2025` = previous_year_over_52,
    `Over 52 weeks: March 2026` = over_52_weeks,
    `Change in count` = annual_change_over_52,
    `Change in proportion (pp)` = annual_change_52wk_pp
  )

knitr::kable(
  long_wait_comparison,
  digits = 2,
  caption = "Pathways waiting over 52 weeks: March 2025 to March 2026"
)
```

### Interpretation

Between March 2025 and March 2026, the number and proportion of
pathways waiting over 52 weeks decreased in every selected
trust–specialty combination.

Imperial College Healthcare recorded the largest absolute reduction
in general surgery, from 121 to 2 pathways. Barts Health recorded
the largest absolute reduction in trauma and orthopaedics, from
181 to 72 pathways.

Guy’s and St Thomas’ recorded a smaller reduction in orthopaedics,
from 86 to 80 pathways. In March 2026, this remained the highest
long-wait count and proportion among the three selected trusts
for that specialty.

These are changes in month-end snapshots. They do not establish
whether pathways left the waiting list through treatment, other
removals or reporting changes.

## 8.3 Summary Heatmap of Annual Changes

These heatmaps summarise changes between March 2025 and March 2026
across the selected trusts and specialties. Labels show the exact
change, rounded to two decimal places.

Each panel uses its own colour scale. Waiting-list changes are
expressed as percentages; changes in waiting-time proportions
are expressed in percentage points (pp).

```{r annual-summary-heatmaps, fig.width=10, fig.height=4.5}
# Select the latest reporting month.
heatmap_data <- rtt_analysis %>%
  filter(reporting_month == max(reporting_month)) %>%
  mutate(
    trust = factor(
      provider_code,
      levels = c("RYJ", "RJ1", "R1H"),
      labels = c(
        "Imperial College Healthcare",
        "Guy's and St Thomas'",
        "Barts Health"
      )
    ),
    specialty = factor(
      specialty_code,
      levels = c("C_100", "C_110", "C_130"),
      labels = c(
        "General\nsurgery",
        "Trauma and\northopaedics",
        "Ophthalmology"
      )
    )
  )

# Create one heatmap for a specified measure.
make_change_heatmap <- function(data, measure, title, unit,
                                higher_is_favourable = FALSE) {

  plot_data <- data %>%
    mutate(
      change = .data[[measure]],
      cell_label = sprintf("%+.2f", change)
    )

  # Centre the colour scale on zero with symmetric limits.
  colour_limit <- max(abs(plot_data$change), na.rm = TRUE)

  if (colour_limit == 0) {
    colour_limit <- 1
  }

  # For within-18-week performance, increases are favourable.
  # For waiting-list size and over-52-week share, decreases
  # are treated as the favourable direction.
  low_colour <- if (higher_is_favourable) "#D55E00" else "#0072B2"
  high_colour <- if (higher_is_favourable) "#0072B2" else "#D55E00"

  ggplot(plot_data, aes(x = specialty, y = trust, fill = change)) +
    geom_tile(colour = "white", linewidth = 1) +
    geom_text(
      aes(label = cell_label),
      colour = "black",
      size = 4.5,
      fontface = "bold"
    ) +
    scale_fill_gradient2(
      low = low_colour,
      mid = "white",
      high = high_colour,
      midpoint = 0,
      limits = c(-colour_limit, colour_limit),
      name = unit
    ) +
    labs(
      title = title,
      subtitle = "March 2026 compared with March 2025",
      x = NULL,
      y = NULL,
      caption = paste(
        "Blue: favourable direction | Orange: adverse direction",
        "Separate colour scale for each panel.",
        sep = "\n"
      )
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      axis.text = element_text(colour = "black"),
      plot.title = element_text(face = "bold"),
      plot.caption = element_text(hjust = 0),
      legend.position = "right"
    )
}

# Print each panel separately so labels remain readable.
print(make_change_heatmap(
  heatmap_data,
  measure = "annual_change_pct",
  title = "A. Change in waiting-list size",
  unit = "Change (%)"
))

print(make_change_heatmap(
  heatmap_data,
  measure = "annual_change_18wk_pp",
  title = "B. Change within 18 weeks",
  unit = "Change (pp)",
  higher_is_favourable = TRUE
))

print(make_change_heatmap(
  heatmap_data,
  measure = "annual_change_52wk_pp",
  title = "C. Change over 52 weeks",
  unit = "Change (pp)"
))
```

### Interpretation

The heatmaps show different patterns across the three indicators.
Waiting-list size increased in five of the nine combinations,
while the proportion within 18 weeks improved in seven.
The proportion waiting over 52 weeks decreased in all nine.

General surgery at Guy’s and St Thomas’ and ophthalmology at
Barts Health combined waiting-list growth with a declining
proportion within 18 weeks. Imperial College Healthcare recorded
smaller lists and improvements in both waiting-time proportions
across all three selected specialties.

Colour intensity should be compared within each panel, rather
than across panels, because the measures have separate scales.
A smaller waiting list is treated here as a favourable direction,
but its causes require investigation: reductions may reflect
treatment, other removals or changes in referrals or reporting.

# 9. Key Findings and Recommendations

## 9.1 Key Findings

### Long waits reduced across all selected services

All nine trust–specialty combinations recorded fewer pathways
waiting over 52 weeks in March 2026 than in March 2025. The
proportion waiting over 52 weeks also fell in every combination.
However, improvements in very long waits did not always coincide
with improvements in the proportion within 18 weeks.

### General surgery at Guy’s and St Thomas’ warrants investigation

The general surgery waiting list increased from 782 to 1,486
pathways between March 2025 and March 2026, an increase of 90.03%.
Over the same period, the proportion within 18 weeks fell by
16.34 percentage points.

The large percentage increase should be interpreted alongside
the relatively small March 2025 baseline. Although pathways
waiting over 52 weeks decreased from 15 to 8, the broader
waiting-time measures indicate a need for further investigation.

### Barts Health ophthalmology showed growing waiting-list pressure

The ophthalmology waiting list increased by 1,000 pathways
(21.82%) between March 2025 and March 2026. The proportion
within 18 weeks fell by 4.79 percentage points, despite a reduction
in pathways waiting over 52 weeks from 18 to 3.

This combination illustrates why monitoring very long waits
alone would provide an incomplete picture of service performance.

### Orthopaedics presents different priorities across trusts

Barts Health had the largest orthopaedic waiting list in March
2026, with 8,564 pathways, and the lowest proportion within
18 weeks among the selected trusts, at 55.73%.

Guy’s and St Thomas’ had a higher proportion within 18 weeks,
at 65.04%, but also the highest count and proportion waiting
over 52 weeks: 80 pathways and 1.64%. These results suggest
different areas for operational review.

### Imperial College Healthcare showed consistent annual improvement

Across all three selected specialties, Imperial College Healthcare
recorded smaller waiting lists, higher proportions within 18 weeks
and fewer pathways waiting over 52 weeks in March 2026 than
in March 2025.

These findings describe observed changes and do not establish
their causes.

## 9.2 Recommendations for Further Investigation

1. **Investigate general surgery at Guy’s and St Thomas’.**
   Review monthly referrals, completed pathways, other removals
   and reporting changes to understand the waiting-list growth
   and decline in the proportion within 18 weeks.

2. **Review ophthalmology demand and activity at Barts Health.**
   Compare additions to the waiting list with pathway completions
   to investigate the growing list and lower proportion within
   18 weeks.

3. **Review orthopaedic waiting-time bands.**
   At Barts Health, examine the distribution of pathways beyond
   18 weeks. At Guy’s and St Thomas’, investigate the remaining
   pathways waiting over 52 weeks.

4. **Monitor several indicators together.**
   Track total pathways, the proportion within 18 weeks, median
   waiting time, and both the count and proportion waiting over
   52 weeks. Interpret these alongside monthly and annual changes.

5. **Validate explanations with operational teams.**
   Discuss the findings with service managers and analysts before
   attributing changes to capacity, demand or service improvements.
   Aggregate RTT snapshots cannot determine the causes on their own.
   
# 10. Limitations and Conclusion

## 10.1 Limitations

- **Selected services only.** The analysis covers three specialties
  at three trusts. Findings cannot be generalised to whole-trust
  performance or the NHS nationally.

- **Pathways rather than unique patients.** A patient may have more
  than one referral-to-treatment pathway. Waiting-list counts
  therefore do not represent unique people.

- **Monthly snapshots.** Each observation represents pathways
  waiting at the end of a reporting month. Adding counts across
  months would count some pathways repeatedly.

- **Descriptive comparisons.** Differences between trusts are not
  adjusted for case complexity, referral patterns, population needs
  or service provision. They should not be treated as an overall
  ranking of care quality.

- **Causes cannot be established.** These data do not explain
  whether changes resulted from referrals, treatment activity,
  other pathway removals or reporting changes.

- **Published data may be revised.** Results reflect the downloaded
  workbooks used in this project. Subsequent NHS England revisions
  may change historical values. Source filenames are retained
  to support traceability.

- **Limited observation period.** Twenty-four months support
  descriptive trend analysis but provide limited evidence about
  longer-term patterns or recurring seasonality.

- **Median waits do not describe the entire distribution.**
  Median waiting times are interpreted alongside the 18-week and
  52-week measures. Specialty medians are not averaged to estimate
  an overall trust median.

## 10.2 Conclusion

This project analysed 216 monthly trust–specialty observations
between April 2024 and March 2026. Checks identified no missing
expected records, duplicate combinations, missing values in the
selected variables or numerical inconsistencies under the
validation rules applied.

Between March 2025 and March 2026, both the number and proportion
of pathways waiting over 52 weeks fell in all nine selected
trust–specialty combinations. However, other indicators showed
a mixed picture. General surgery at Guy’s and St Thomas’ and
ophthalmology at Barts Health combined waiting-list growth with
a declining proportion of pathways within 18 weeks.

The findings demonstrate the value of examining waiting-list size,
typical waiting time and long waits together. They identify areas
for further investigation rather than establish the causes of
performance changes.

# 11. Sources and Reproducibility

## 11.1 Data Sources

The analysis uses NHS England's monthly incomplete
referral-to-treatment provider workbooks, using the `Provider`
worksheet.

- [NHS England RTT data: 2024–25](https://www.england.nhs.uk/statistics/statistical-work-areas/rtt-waiting-times/rtt-data-2024-25/)
- [NHS England RTT data: 2025–26](https://www.england.nhs.uk/statistics/statistical-work-areas/rtt-waiting-times/rtt-data-2025-26/)

The selected reporting months run from April 2024 to March 2026.
Original workbooks are retained in the project's `data` folder.

## 11.2 Source File Record

The following table links each reporting month to the workbook
used in the analysis.

```{r source-file-record}
source_file_record <- rtt_data %>%
  distinct(reporting_month, source_file) %>%
  arrange(reporting_month)

knitr::kable(
  source_file_record,
  caption = "Source workbooks used for each reporting month"
)
```

## 11.3 Reproducing the Report

To reproduce the analysis:

1. Open the RStudio project.
2. Ensure the 24 source workbooks are in the `data` folder.
3. Install the required packages: `readxl`, `dplyr`, `tidyr`,
   `ggplot2`, `knitr` and `rmarkdown`.
4. Knit this R Markdown file to HTML.

The report imports the workbooks, selects the project scope,
validates the data, calculates indicators and generates the
tables and charts.

## 11.4 Software Environment

The output below records the R version, platform and loaded
package versions used when rendering this report.

```{r session-information}
sessionInfo()
```


































































































































